---
title: "DR_Analysis_Main"
author: "Vikas Maturi"
date: 2021-07-07
output: 
  github_document:
    md_extension: +gfm_auto_identifiers
    toc: true
    toc_depth: 6
---


## Setup

#### Libraries and data import

```{r, results=FALSE, message=FALSE}

# load libraries
library(tidyverse)
library(readxl)
library(lubridate)
library(knitr)
library(googlesheets4)
library(car)


# parameters

race_colors <- c("#54AE55", "#189CD9", "#A167A5")

insurance_colors <- c("#A00033", "#D1603D", "#F7D002")

```

```{r, results = FALSE, message = FALSE, warning = FALSE}

# read in cleaned, combined data on visits and severity
visits_severity <- read_csv(file = "data_clean/maturi_all_cleaned.csv")

# read in cleaned data on timeservies visual acuity
va_progression <- read_csv(file = "data_clean/va_progression.csv")

```

#### Join severity and timeseries patient/visit data

```{r}

joined_data <- 
  visits_severity %>% 
  # select the data
  dplyr::select(pt_code_name, eye, first_problem_code, severity_score, gender, race_ethnicity, first_dr_age, age_group, insurance, region, smoke_status, new_class, valid_class, include_given_code, vision_category, vision_threatening, pt_code_name, eye, index_date, baseline_va_letter, proc_group_28, proc_group_365, proc_group_any, vegf_group_28, vegf_group_365, vegf_group_any, retina_speciality, baseline_iop, pdr_group, cat_eyes) %>% 
  left_join(va_progression, by = c("pt_code_name" = "patient_guid", "eye" = "va_eye", "index_date"))

```


## Cleaned datasets for analysis

The 'timeseries_analysis' dataset will be the basis of all analyses
```{r}

# set seed to ensure that we randomly sample the same set of folks in each round
set.seed(1)
# filter out the necesssary data to conduct time series analysis
timeseries_analysis <-
  joined_data %>% 
  filter(
    # only data classified under new ICD codes or codes starting of the format 36x.xxx
    include_given_code == 1,
    #race is White, Black, or Hispanic, to ensure adequate numbers
    race_ethnicity %in% c("Caucasian", "Black or African American", "Hispanic"),
    #race and gender data available for the patient
    !is.na(race_ethnicity), 
    !is.na(gender),
    !is.na(severity_score),
    # only patients with one year and two year data available
    !is.na(one_year), 
    #!is.na(six_month),
    !is.na(two_year),
    # only patients getting antivegf treatment
    proc_group_28 == "antivegf",
    # no cat eyes
    !cat_eyes == 1
  ) %>% 
  # select only one eye per patient
  group_by(pt_code_name) %>%
  sample_n(1) %>%
  ungroup() %>%
  # adjust insurance data to combine medicare FFS and medicare managed
  mutate(
    insurance = if_else(insurance %in% c("Medicare FFS", "Medicare Managed"), "Medicare", insurance)
  ) %>% 
  mutate(
    baseline_va_r10 = plyr::round_any(baseline_va_letter, 10),
    baseline_va_quart = ntile(baseline_va_letter, 4),
    one_year_va_delta = one_year - baseline_va_letter,
    two_year_va_delta = two_year - baseline_va_letter,
    #turn race into a factor and rename races
    race_ethnicity = dplyr::case_when(
      race_ethnicity == "Caucasian" ~ "White",
      race_ethnicity == "Black or African American" ~ "Black",
      TRUE ~ race_ethnicity
    ),
    race_ethnicity = factor(race_ethnicity, ordered = TRUE, levels = c("White", "Black", "Hispanic")),
    insurance = factor(insurance, ordered = TRUE, levels = c("Private", "Medicare","Medicaid","Govt","Military","Unknown/Missing"))
  ) 

```


## Table 1: Patient Demographics

#### Construct dataset to generate data tables

```{r}

# Create dataset for basic paper tables
tbl_data <-
  timeseries_analysis %>% 
  mutate(
    # unorder insurance to permit grouping
    insurance = as.character(insurance),
    insurance_tbl = if_else(
      insurance %in% c("Private", "Medicare", "Medicaid"),
      insurance, 
      "Other"
    ),
    age_tbl = case_when(
      first_dr_age < 53 ~ "< 53",
      first_dr_age >= 53 & first_dr_age <= 58 ~ "53 - 58",
      first_dr_age >= 59 & first_dr_age <= 64 ~ "59 - 64",
      first_dr_age >= 65 & first_dr_age <= 70 ~ "65 - 70",
      first_dr_age > 70 ~ "> 70"
    ),
    va_tbl = case_when(
      baseline_va_letter > 69.95 ~ "20/40 or better",
      baseline_va_letter >= 57.80 & baseline_va_letter <= 69.95 ~ "20/41 - 20/70",
      baseline_va_letter >= 35 & baseline_va_letter <= 57.80 ~ "20/71 - 20/200",
      baseline_va_letter < 35 ~ "20/201 or worse"
    ), 
    vegf_tbl = case_when(
      vegf_group_365 == "Avastin" ~ "bevacizumab" ,
      vegf_group_365 == "Lucentis" ~ "ranibizumab",
      vegf_group_365 == "Eylea" ~ "aflibercept",
      vegf_group_365 == "combo" ~ "combo"
    )
  )

tbl_data
```


#### Table: Age by race and insurance

```{r}
tbl_age_race <-
  tbl_data %>% 
  count(age_tbl, race_ethnicity) %>% 
  group_by(race_ethnicity) %>% 
  mutate(pct_age_race = round(n / sum(n) * 100, 3)) %>% 
  unite(x, "n", "pct_age_race", sep = " (") %>% 
  spread(key = race_ethnicity, value = x )

kable(tbl_age_race)


tbl_age_insurance <-
  tbl_data %>% 
  count(age_tbl, insurance_tbl) %>% 
  group_by(insurance_tbl) %>% 
  mutate(pct_age_insurance = round(n / sum(n) * 100, 3)) %>%
  unite(x, "n", "pct_age_insurance", sep = " (") %>% 
  spread(key = insurance_tbl, value = x )

kable(tbl_age_insurance)

```


#### Table: Gender by race and insurance

```{r}

tbl_gender_race <-
  tbl_data %>% 
  count(gender, race_ethnicity) %>% 
  group_by(race_ethnicity) %>% 
  mutate(pct_age_race = round(n / sum(n) * 100, 3)) %>% 
  unite(x, "n", "pct_age_race", sep = " (") %>% 
  spread(key = race_ethnicity, value = x )

kable(tbl_gender_race)

tbl_gender_insurance <-
  tbl_data %>% 
  count(gender, insurance_tbl) %>% 
  group_by(insurance_tbl) %>% 
  mutate(pct_gender_insurance = round(n / sum(n) * 100, 3)) %>%
  unite(x, "n", "pct_gender_insurance", sep = " (") %>% 
  spread(key = insurance_tbl, value = x )

kable(tbl_gender_insurance)
```


#### Table: Baseline VA by race and insurance

```{r}
tbl_va_race <-
  tbl_data %>% 
  count(va_tbl, race_ethnicity) %>% 
  group_by(race_ethnicity) %>% 
  mutate(pct_va_race = round(n / sum(n) * 100, 3)) %>% 
  unite(x, "n", "pct_va_race", sep = " (") %>% 
  spread(key = race_ethnicity, value = x )

kable(tbl_va_race)


tbl_va_insurance <-
  tbl_data %>% 
  count(va_tbl, insurance_tbl) %>% 
  group_by(insurance_tbl) %>% 
  mutate(pct_va_insurance = round(n / sum(n) * 100, 3)) %>%
  unite(x, "n", "pct_va_insurance", sep = " (") %>% 
  spread(key = insurance_tbl, value = x )

kable(tbl_va_insurance)
```


#### Table: VEGF drug by race and insurance

```{r}
tbl_vegf_race <-
  tbl_data %>% 
  count(vegf_tbl, race_ethnicity) %>% 
  group_by(race_ethnicity) %>% 
  mutate(pct_vegf_race = round(n / sum(n) * 100, 3)) %>% 
  unite(x, "n", "pct_vegf_race", sep = " (") %>% 
  spread(key = race_ethnicity, value = x )

kable(tbl_vegf_race)

tbl_vegf_insurance <-
  tbl_data %>% 
  count(vegf_tbl, insurance_tbl) %>% 
  group_by(insurance_tbl) %>% 
  mutate(pct_vegf_insurance = round(n / sum(n) * 100, 3)) %>%
  unite(x, "n", "pct_vegf_insurance", sep = " (") %>% 
  spread(key = insurance_tbl, value = x )

kable(tbl_vegf_insurance)
```


#### Table: Smoke status by race and insurance

```{r}
tbl_smoke_race <-
  tbl_data %>% 
  count(smoke_status, race_ethnicity) %>% 
  group_by(race_ethnicity) %>% 
  mutate(pct_smoke_race = round(n / sum(n) * 100, 3)) %>% 
  unite(x, "n", "pct_smoke_race", sep = " (") %>% 
  spread(key = race_ethnicity, value = x )


kable(tbl_smoke_race)

tbl_smoke_insurance <-
  tbl_data %>% 
  count(smoke_status, insurance_tbl) %>% 
  group_by(insurance_tbl) %>% 
  mutate(pct_smoke_insurance = round(n / sum(n) * 100, 3)) %>%
  unite(x, "n", "pct_smoke_insurance", sep = " (") %>% 
  spread(key = insurance_tbl, value = x )

kable(tbl_smoke_insurance)
```


#### Table: Insurance by race

```{r}
tbl_insurance_race <-
  tbl_data %>% 
  count(insurance_tbl, race_ethnicity) %>% 
  group_by(race_ethnicity) %>% 
  mutate(pct_insurance_race = n / sum(n) * 100)

kable(tbl_insurance_race)
```


## Calculation X: Mean VA and SD of VA and at baseline

```{r}

mean_va_race <-
  tbl_data %>% 
  group_by(race_ethnicity) %>% 
  summarize(
    mean_baseline_va_letter = mean(baseline_va_letter),
    sd = sd(baseline_va_letter)
  )

kable(mean_va_race)


mean_va_insurance <-
  tbl_data %>% 
  group_by(insurance) %>% 
  summarize(
    mean_baseline_va_letter = mean(baseline_va_letter),
    sd = sd(baseline_va_letter),
    n = n()
  )

kable(mean_va_insurance)
```


## Calculation X: N after eliminating ICD-10 diagnostic codes

```{r}

tbl_data %>% 
  filter(!is.na(vision_category)) %>% 
  count()
  
```



## Figure 1: Baseline Differences in Severity Score by Race

To consider here: how do we want to respond to the NAs in this graph? One option I could see is creating four new categories: (mild NPDR unknown edema, moderate NPDR unknown edema, severe NPDR unknown edema, PDR unknown edema)

```{r}
timeseries_analysis %>% 
  dplyr::count(vision_category, race_ethnicity) %>%
  dplyr::ungroup() %>% 
  dplyr::group_by(race_ethnicity) %>% 
  dplyr::mutate(prop_category = n / sum(n)) %>% 
  ggplot() + 
  geom_col(aes(x = vision_category, y = prop_category, group = race_ethnicity, fill = race_ethnicity), position = "dodge", alpha = 1) +
  theme_light() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 10)) +
  scale_fill_manual(values = race_colors) +
  labs(
    x = "Severity Category",
    y = "Percentage ",
    fill = "Race"
  )
```


## Figure 2A and 2B: Differences in Rates of Bevacizumab Injection by Severity Category, Race and Insurance Status

```{r}

# Calculate likelihood of drug by severity score and race (given that patient received antivegf treatment)
drug_sev_cat_race <-
  timeseries_analysis %>% 
  filter(proc_group_365 == "antivegf") %>% 
  count(vision_category, race_ethnicity, vegf_group_365) %>% 
  group_by(vision_category, race_ethnicity) %>% 
  mutate(count_severity_race = sum(n)) %>% 
  mutate(prop = n / count_severity_race) %>% 
  ungroup() %>% 
  select(vision_category, race_ethnicity, vegf_group_365, prop, count = n)

drug_sev_cat_race %>% 
  arrange(desc(vision_category), race_ethnicity) #JAY changed this!! 

```

```{r}

# Graphing likelihood of Bevacizumab by severity and race
plot_drug_sev_cat_race <-
  drug_sev_cat_race %>% 
  # remove individuals with no vision category
  filter(!is.na(vision_category), !vision_category %in% c("NA")) %>% 
  # remove combo, Eylea, and Lucentis for this graph
  filter(vegf_group_365 == "Avastin") %>% 
  # round proportion to nearest 100th
  mutate(pct_round = round(prop, 2)) %>% 
  ggplot(aes(x = vision_category, y = pct_round, fill = race_ethnicity)) +
  geom_col(position = "dodge") +
  theme_light() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  #facet_grid(vegf_group_365 ~ .) +
  #geom_text(aes(label = paste0(pct_round, "%")), vjust = -0.5, position = position_dodge(width = 1)) +
  #scale_x_continuous()
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), limits = c(0, 1)) +
  scale_fill_manual(values = race_colors) +
  labs(
    x = "Severity category",
    y = "Percentage of eyes receiving Bevacizumab",
    fill = "Race"
  )

plot_drug_sev_cat_race

#ggsave("graphs/2A_plot_drug_sev_cat_race.png", plot_drug_sev_cat_race, width = 10, height = 5)

```


```{r}

# Calculate likelihood of drug by severity score and insurance (given that patient received antivegf treatment)
drug_sev_cat_insurance <-
  timeseries_analysis %>%
  filter(proc_group_365 == "antivegf", insurance %in% c("Private", "Medicare", "Medicaid")) %>% 
  count(vision_category, insurance, vegf_group_365) %>% 
  group_by(vision_category, insurance) %>% 
  mutate(count_severity_race = sum(n)) %>% 
  mutate(prop = n / count_severity_race) %>% 
  ungroup() %>% 
  select(vision_category, insurance, vegf_group_365, prop, count = n)

drug_sev_cat_insurance %>% 
  arrange(desc(vision_category), insurance)


```

```{r}

plot_drug_sev_cat_insurance <-
# Graphing likelihood of first drug type by severity and insurance
drug_sev_cat_insurance %>% 
  # remove individuals with no vision category
  filter(!is.na(vision_category), !vision_category %in% c("NA")) %>% 
  # remove combo, Eylea, and Lucentis for this graph
  filter(vegf_group_365 == "Avastin") %>% 
  # round proportion to nearest 100th
  mutate(pct_round = round(prop, 2)) %>%
  ggplot(aes(x = vision_category, y = prop, fill = insurance)) +
  geom_col(position = "dodge") +
  theme_light() +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1),
    # remove background lines
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  #facet_grid(vegf_group_365~.) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), limits = c(0, 1)) +
  # different colors for insurance
  scale_fill_manual(values = insurance_colors) +
  labs(
    x = "Severity category",
    y = "Percentage of people receiving Bevacizumab",
    fill = "Insurance"
  )

plot_drug_sev_cat_insurance

#ggsave("graphs/2B_plot_drug_sev_cat_insurance.png", plot_drug_sev_cat_insurance, width = 10, height = 5)


```

## Regression X: Variables that influence likelihood of receiving Bevacizumab treatment

```{r}

# create dataset for regression
data_avastin_reg <- 
  timeseries_analysis %>% 
  filter(
    insurance %in% c("Medicare", "Medicaid", "Private")
  ) %>% 
  mutate(
    # unorder for appropriate regression interpretation
    race_ethnicity = factor(race_ethnicity, ordered = FALSE),
    race_ethnicity = relevel(race_ethnicity, ref = "White"),
    # unorder for appropriate regression interpretation
    insurance = factor(insurance, ordered = FALSE),
    insurance = relevel(insurance, ref = "Private"),
    smoke_status = factor(smoke_status),
    smoke_status = relevel(smoke_status, ref = "No / Never"),
    avastin = if_else(vegf_group_365 == "Avastin", 1, 0)
  ) 

# glm for logistic regression
avastin_1.logm <- 
  glm(
    avastin ~ race_ethnicity + insurance + race_ethnicity * insurance + gender + first_dr_age + smoke_status + baseline_va_letter +  severity_score, 
    data = data_avastin_reg, 
    family = binomial
  ) 

summary(avastin_1.logm)

```

## Supplement: Differences in Rates of Bevacizumab, Aflibercept, and Ranibizumab Injection by Severity Category, Race and Insurance Status

```{r}

plot_drug_sev_cat_race_all <- 
# Graphing likelihood of Bevacizumab by severity and race
drug_sev_cat_race %>% 
  # remove individuals with no vision category
  filter(!is.na(vision_category), !vision_category %in% c("NA")) %>% 
  # round proportion to nearest 100th
  mutate(pct_round = round(prop, 2)) %>% 
  ggplot(aes(x = vision_category, y = pct_round, fill = race_ethnicity)) +
  geom_col(position = "dodge") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1),
    # remove background lines
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +  facet_grid(vegf_group_365 ~ .) +
  #geom_text(aes(label = paste0(pct_round, "%")), vjust = -0.5, position = position_dodge(width = 1)) +
  #scale_x_continuous()
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), limits = c(0, 1)) +
  scale_fill_manual(values = race_colors) +
  labs(
    x = "Severity category",
    y = "Percentage of people receiving Bevacizumab",
    fill = "Race"
  )

plot_drug_sev_cat_race_all

#ggsave("graphs/SUPPLEMENT_plot_drug_sev_cat_race_all.png", plot_drug_sev_cat_race_all, width = 10, height = 7)


```


```{r}
plot_drug_sev_cat_insurance_all <-
# Graphing likelihood of first drug type by severity and insurance
drug_sev_cat_insurance %>% 
  # remove individuals with no vision category
  filter(!is.na(vision_category), !vision_category %in% c("NA")) %>% 
  # round proportion to nearest 100th
  mutate(pct_round = round(prop, 2)) %>%
  ggplot(aes(x = vision_category, y = prop, fill = insurance)) +
  geom_col(position = "dodge") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1),
    # remove background lines
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +  facet_grid(vegf_group_365 ~ .) +
  facet_grid(vegf_group_365~.) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), limits = c(0, 1)) +
  # different colors for insurance
  scale_fill_manual(values = insurance_colors) +
  labs(
    x = "Severity category",
    y = "Percentage of people receiving Bevacizumab",
    fill = "Insurance"
  )

plot_drug_sev_cat_insurance_all

#ggsave("graphs/SUPPLEMENT_plot_drug_sev_cat_insurance_all.png", plot_drug_sev_cat_insurance_all, width = 10, height = 7)

```




## Figure 3A and 3B: Visual Acuity Change from Baseline at One and Two Years, by Race

```{r}

change_15 <- 
  timeseries_analysis %>% 
  mutate(
    one_year_change = one_year - baseline_va_letter,
    two_year_change = two_year - baseline_va_letter,
    baseline_change = 0
  ) %>% 
  mutate(
    one_year_15_pt_loss = if_else(one_year_change <= -15, 1, 0),
    two_year_15_pt_loss = if_else(two_year_change <= -15, 1, 0),
    one_year_15_pt_gain = if_else(one_year_change >= 15, 1, 0),
    two_year_15_pt_gain = if_else(two_year_change >= 15, 1, 0),
  ) %>% 
  #gather(key = "time_diff", value = "va_change", baseline_change, one_year_change, two_year_change) %>% 
  ungroup()

# z value for p = .05 
z = 1.96

```

```{r}

loss_15_race <-
  change_15 %>% 
  group_by(race_ethnicity) %>% 
  summarize(
    one_year_loss = sum(one_year_15_pt_loss) / n(),
    two_year_loss = sum(two_year_15_pt_loss) / n(),
    one_year_sd = sqrt(one_year_loss * (1 - one_year_loss) / n()),
    two_year_sd = sqrt(two_year_loss * (1 - two_year_loss) / n())
  ) %>% 
  gather(key = "time_diff", value = "pct_15pt_loss", one_year_loss, two_year_loss) %>% 
  mutate(
    sd = if_else(str_detect(string = time_diff, pattern = "one_year"), one_year_sd, two_year_sd),
    upper_bound = pct_15pt_loss + z * sd,
    lower_bound = pct_15pt_loss - z * sd
  ) %>%
  mutate(
    time_diff = case_when(
      time_diff == "one_year_loss" ~ "One year",
      time_diff == "two_year_loss" ~ "Two year"
    )
  )

plot_loss_15_race <-
loss_15_race %>%
  ggplot(aes(x = time_diff, y = pct_15pt_loss, fill = race_ethnicity)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), position = position_dodge(width = .9), width = .3) +
  theme_light() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  scale_fill_manual(values = race_colors) +
  labs(
    x = "Time after baseline VA measurement",
    y = "Percentage of eyes that lost 15 or more pts of VA",
    fill = "Race"
  )

plot_loss_15_race

ggsave("graphs/3A_plot_loss_15_race.png", plot_loss_15_race, width = 7, height = 5)


```

```{r}
# plot change in visual acuity by race at one and two years

va_race <-
  timeseries_analysis %>% 
  group_by(race_ethnicity) %>% 
  mutate(
    one_year_change_new = one_year - baseline_va_letter,
    two_year_change_new = two_year - baseline_va_letter
  ) %>% 
  summarize(
    baseline_va = mean(baseline_va_letter), 
    one_year_va = mean(one_year), 
    two_year_va = mean(two_year),
    baseline_sd = sd(baseline_va_letter),
    one_year_sd = sd(one_year),
    two_year_sd = sd(two_year),
    one_year_change_sd = sd(one_year_change_new),
    two_year_change_sd= sd(two_year_change_new),
    count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()

va_race_diff <-
  va_race %>% 
  spread(key = timepoint, value = va_plot) %>% 
  mutate(
    one_year_change = one_year_va - baseline_va,
    two_year_change = two_year_va - baseline_va,
    baseline = 0
  ) %>% 
  gather(key = "time_diff", value = "va_change", baseline, one_year_change, two_year_change) %>% 
  mutate(
    time_diff = case_when(
      time_diff == "baseline" ~ "Baseline",
      time_diff == "one_year_change" ~ "One year",
      time_diff == "two_year_change" ~ "Two years"
    )
  ) %>% 
  ungroup()

plot_va_race_diff <- 
  va_race_diff %>% 
  ggplot(aes(x = time_diff, y = va_change, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = race_ethnicity)) +
  # ggrepel::geom_text_repel(
  #   data = va_race_diff %>% filter(time_diff == "two_year_change"),
  #   aes(label = count)
  # ) +
  theme_light() +
  theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom",
    #legend.spacing.x = unit(1.0, "cm")
    legend.margin = margin(c(5, 5, 5, 0)),
    legend.text = element_text(margin = margin(r = 10, unit = "pt"))
  ) +
  scale_color_manual(values = race_colors) +
  labs(
    x = "",
    y = "Change in visual acuity (VA)",
    color = "Race"
  ) 
  # guides(
  #   fill = guide_legend(
  #     label.position = "top",
  #     #title.position = "left", 
  #   )
  # )

plot_va_race_diff

#ggsave("graphs/3B_plot_va_race_diff.png", plot_va_race_diff, width = 7, height = 5)
```

## Figure 4A, 4B, 4C: Visual Acuity Change from Baseline at One and Two Years, by Insurance and Race

```{r}

# z value for p = .05 
z = 1.96

loss_15_insurance <-
  change_15 %>% 
  group_by(insurance) %>%
  filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>% 
  summarize(
    one_year_loss = sum(one_year_15_pt_loss) / n(),
    two_year_loss = sum(two_year_15_pt_loss) / n(),
    one_year_sd = sqrt(one_year_loss * (1 - one_year_loss) / n()),
    two_year_sd = sqrt(two_year_loss * (1 - two_year_loss) / n())
  ) %>% 
  gather(key = "time_diff", value = "pct_15pt_loss", one_year_loss, two_year_loss) %>% 
  mutate(
    sd = if_else(str_detect(string = time_diff, pattern = "one_year"), one_year_sd, two_year_sd),
    upper_bound = pct_15pt_loss + z * sd,
    lower_bound = pct_15pt_loss - z * sd
  ) %>% 
  mutate(
    time_diff = case_when(
      time_diff == "one_year_loss" ~ "One year",
      time_diff == "two_year_loss" ~ "Two year"
    )
  )

plot_loss_15_insurance <-
loss_15_insurance %>% 
  ggplot(aes(x = time_diff, y = pct_15pt_loss, fill = insurance)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), position = position_dodge(width = .9), width = .4) +
  theme_light() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  scale_fill_manual(values = insurance_colors) +
  labs(
    x = "Time after baseline VA measurement",
    y = "Percentage of eyes that lost 15 or more points of VA",
    fill = "Insurance"
  )

plot_loss_15_insurance

ggsave("graphs/4A_plot_loss_15_insurance.png", plot_loss_15_insurance, width = 7, height = 5)
```

```{r}

va_insurance <-
  timeseries_analysis %>% 
  group_by(insurance) %>%
  filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>% 
  summarize(
    baseline_va = mean(baseline_va_letter), 
    one_year_va = mean(one_year), 
    two_year_va = mean(two_year),
    baseline_sd = sd(baseline_va_letter),
    one_year_sd = sd(one_year),
    two_year_sd = sd(two_year),
    count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()

va_insurance_diff <-
  va_insurance %>% 
  spread(key = timepoint, value = va_plot) %>% 
  mutate(
    one_year_change = one_year_va - baseline_va,
    two_year_change = two_year_va - baseline_va,
    baseline = 0
  ) %>% 
  gather(key = "time_diff", value = "va_change", baseline, one_year_change, two_year_change) %>% 
  mutate(
    time_diff = case_when(
      time_diff == "baseline" ~ "Baseline",
      time_diff == "one_year_change" ~ "One year",
      time_diff == "two_year_change" ~ "Two years"
    )
  ) %>% 
  ungroup()

plot_va_insurance_diff <-
va_insurance_diff %>% 
  ggplot(aes(x = time_diff, y = va_change, color = insurance)) +
  geom_point() +
  geom_line(aes(group = insurance)) +
  # ggrepel::geom_text_repel(
  #   data = va_insurance_diff %>% filter(time_diff == "two_year_change"),
  #   aes(label = count)
  # ) +
  theme_light() + 
  theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  scale_color_manual(values = insurance_colors) +
  labs(
    x = "Time point",
    y = "Change in visual acuity",
    color = "Insurance"
  )

plot_va_insurance_diff

#ggsave("graphs/4B_plot_va_insurance_diff.png", plot_va_insurance_diff, width = 7, height = 5)

```


```{r}

insurance_race_va <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, insurance) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va)

va_insurance_race_diff_tbl <-
  insurance_race_va %>%
  filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>%
  spread(key = timepoint, value = va_plot) %>%
  mutate(
    one_year_change = one_year_va - baseline_va,
    two_year_change = two_year_va - baseline_va,
    baseline = 0
  )

va_insurance_race_diff <-
  insurance_race_va %>%
  filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>%
  spread(key = timepoint, value = va_plot) %>%
  mutate(
    one_year_change = one_year_va - baseline_va,
    two_year_change = two_year_va - baseline_va,
    baseline = 0
  ) %>%
  gather(key = "time_diff", value = "va_change", baseline, one_year_change, two_year_change) %>%
  mutate(
    time_diff = case_when(
      time_diff == "baseline" ~ "Baseline",
      time_diff == "one_year_change" ~ "One year",
      time_diff == "two_year_change" ~ "Two years"
    )
  ) %>% 
  ungroup()

va_insurance_race_diff_ci <-
  timeseries_analysis %>%
  mutate(
    one_year_change = one_year - baseline_va_letter,
    two_year_change = two_year - baseline_va_letter,
    baseline = 0
  ) %>%
  group_by(race_ethnicity, insurance) %>%
  summarize(
    mean_one_year_change = mean(one_year_change),
    mean_two_year_change = mean(two_year_change),
    sd_one_year_change = sd(one_year_change),
    sd_two_year_change = sd(two_year_change)
  )

plot_va_insurance_race_diff <-
va_insurance_race_diff %>%
  ggplot(aes(x = time_diff, y = va_change, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(insurance, race_ethnicity))) +
  # to display counts
  # ggrepel::geom_text_repel(
  #   data = va_insurance_race_diff %>% filter(time_diff == "two_year_change", insurance %in% c("Private", "Medicaid", "Medicare")),
  #   aes(label = count, hjust = -2)
  # ) +
  facet_grid(.~insurance) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1),
    # remove background lines
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  scale_color_manual(values = race_colors) +
  labs(
    x = "Time point",
    y = "Visual acuity (VA) change",
    color = "Race"
  )

plot_va_insurance_race_diff

#ggsave("graphs/4C_plot_va_insurance_race_diff.png", plot_va_insurance_race_diff, width = 10, height = 5)


```

## Regression X: Variables that influence change in vision after one year

```{r}
# linear regression with race insurance interaction and severity score
one_year_va_delta_4.lm <- 
  lm(one_year_va_delta ~ race_ethnicity + insurance + race_ethnicity * insurance + gender + first_dr_age + smoke_status + baseline_va_letter + vegf_group_365 + severity_score, 
     data = timeseries_analysis %>% 
       filter(
         insurance %in% c("Medicare", "Medicaid", "Private")
       ) %>% 
       mutate(
         race_ethnicity = factor(race_ethnicity, ordered = FALSE),
         race_ethnicity = relevel(race_ethnicity, ref = "White"),
         insurance = factor(insurance, ordered = FALSE), 
         insurance = relevel(insurance, ref = "Private"),
         smoke_status = factor(smoke_status),
         smoke_status = relevel(smoke_status, ref = "No / Never")
       )
  ) 

summary(one_year_va_delta_4.lm)

vif(one_year_va_delta_4.lm)
```

## Calculation X: Number of patients including the multivariate linear regresison

```{r}
timeseries_analysis %>% 
  filter(
    insurance %in% c("Medicare", "Medicaid", "Private")
  ) %>% 
  mutate(
    race_ethnicity = factor(race_ethnicity, ordered = FALSE),
    race_ethnicity = relevel(race_ethnicity, ref = "White"),
    insurance = factor(insurance, ordered = FALSE), 
    insurance = relevel(insurance, ref = "Private"),
    smoke_status = factor(smoke_status),
    smoke_status = relevel(smoke_status, ref = "No / Never")
  ) %>% 
  count()
```


## Figure 5: Average severity score at each baseline VA (rounded to nearest 5) by race

```{r}

#calculate severity by race and age

severity_race_vision <-
  timeseries_analysis %>% 
  mutate(baseline_va_r5 = plyr::round_any(baseline_va_letter, 5)) %>% 
  group_by(race_ethnicity, baseline_va_r5) %>%
  summarize(avg_severity = mean(severity_score), count = n()) %>% 
  filter(count > 100) %>% 
  arrange(desc(avg_severity))

severity_race_vision

```

```{r}
# graph severity by race and age group

plot_sev_race_va <-
  severity_race_vision %>%
  filter(race_ethnicity %in% c("White", "Black", "Hispanic")) %>% 
  ggplot(aes(x = baseline_va_r5, y = avg_severity, color = race_ethnicity)) +
  geom_line() +
  geom_point() +
  theme_light() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  scale_color_manual(values = race_colors) +
  labs(
    y = "Average severity score", 
    x = "Baseline VA", 
    color = "Race"
  )

plot_sev_race_va

#ggsave("graphs/5_plot_sev_race_va.png", plot_sev_race_va, width = 7, height = 5)

```


## Figure 6: Average severity score at each baseline VA (rounded to nearest 5) by insurance

```{r}

#calculate severity by race and age

severity_insurance_vision <-
  timeseries_analysis %>% 
  mutate(baseline_va_r5 = plyr::round_any(baseline_va_letter, 5)) %>% 
  group_by(insurance, baseline_va_r5) %>%
  summarize(avg_severity = mean(severity_score), count = n()) %>% 
  filter(count > 100) %>% 
  arrange(desc(avg_severity))

severity_insurance_vision

```

```{r}
# graph severity by race and age group
plot_sev_insurance_va <-
  severity_insurance_vision %>%
  filter(insurance %in% c("Medicaid", "Medicare", "Private")) %>% 
  ggplot(aes(x = baseline_va_r5, y = avg_severity, color = insurance)) +
  geom_line() +
  geom_point() +
  theme_light() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  scale_color_manual(values = insurance_colors) +
  labs(
    title = "Severity by insurance and baseline VA", 
    y = "Average severity score", 
    x = "Baseline VA", 
    color = "Insurance"
  )

plot_sev_insurance_va

#ggsave("graphs/6_plot_sev_insurance_va.png", plot_sev_insurance_va, width = 7, height = 5)

```




## Timeseries analysis of visual acuity
#### Distribution of VA by race

```{r}
# table of percentage distribution to baseline letter va by race
race_distribution <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, baseline_va_r10) %>%   
  summarize(count = n()) %>% 
  ungroup() %>% 
  group_by(race_ethnicity) %>% 
  mutate(prop = count / sum(count)) %>% 
  ungroup()

race_distribution %>% 
  group_by(race_ethnicity) %>% 
  summarize(total_race = sum(count))
```

```{r}
timeseries_analysis %>% 
  ggplot(aes(x = baseline_va_letter)) +
  geom_histogram(binwidth = 10, aes(fill = race_ethnicity)) +
  theme_light() +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  labs(
    title = "Histogram of eye count by race"
  )

```


```{r}
race_distribution %>% 
  ggplot(aes(x = baseline_va_r10, y = prop, color = race_ethnicity)) + 
  geom_line() +
  geom_point() +
  theme_light()  +
  scale_y_continuous() +
  labs(
    title = "Propotion of eyes in each group of baseline visual acuity, by race",
    y = "Proportion", 
    x = "Baseline VA, rounded to nearest multiple of 10"
  )
```


#### Timeseries VA by race and severity

```{r}
## calculate change in va over time for people in each severity group/race
severity_race_va <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, severity_score) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va)

```


```{r}
timeseries_analysis %>% 
  count(severity_score, race_ethnicity) %>% 
  group_by(race_ethnicity) %>% 
  mutate(prop = n / sum(n))
```


```{r}
## density chart of distribution of severity scores, by race
timeseries_analysis %>% 
  ggplot() +
  geom_density(aes(x = severity_score, group = race_ethnicity, color = race_ethnicity), alpha = .2) +
  theme_light()
```

This chart shows the density of patients by severity score. There are two peaks - first at around severity scores between 40-48, and a second at severity scores betewen 73-78. These numbers are generally aligned to categorizations of moderate NPDR (with and without edema), and PDR (with and without edema).

#### Timeseries VA by race 
```{r}
# plot change in visual acuity by race at one and two yeaars

va_race <-
  timeseries_analysis %>% 
  group_by(race_ethnicity) %>% 
  summarize(
    baseline_va = mean(baseline_va_letter), 
    one_year_va = mean(one_year), 
    two_year_va = mean(two_year),
    baseline_sd = sd(baseline_va_letter),
    one_year_sd = sd(one_year),
    two_year_sd = sd(two_year),
    count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()
  
va_race %>% 
  ggplot(aes(x = timepoint, y = va_plot, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = race_ethnicity)) +
  ggrepel::geom_text_repel(
    data = va_race %>% filter(timepoint == "two_year_va"),
    aes(label = count)
  ) +
  theme_light() + 
  labs(
    title = "VA at baseline, one, and two years by race",
    y = "Visual acuity"
  )

```

```{r}

va_race_diff <-
  va_race %>% 
  spread(key = timepoint, value = va_plot) %>% 
  mutate(
    one_year_change = one_year_va - baseline_va,
    two_year_change = two_year_va - baseline_va,
    baseline = 0
  ) %>% 
  gather(key = "time_diff", value = "va_change", baseline, one_year_change, two_year_change) %>% 
  ungroup()

va_race_diff %>% 
  ggplot(aes(x = time_diff, y = va_change, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = race_ethnicity)) +
  ggrepel::geom_text_repel(
    data = va_race_diff %>% filter(time_diff == "two_year_change"),
    aes(label = count)
  ) +
  theme_light() + 
  labs(
    title = "VA change from baseline, one, and two years by race",
    y = "Visual acuity (VA) change"
  )
```


```{r}
# plot change in visual acuity by race at one and two yeaars
severity_race_va %>% 
  filter(severity_score == 73) %>% 
  ggplot(aes(x = timepoint, y = va_plot, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = race_ethnicity)) +
  ggrepel::geom_text_repel(
    data = severity_race_va %>% filter(timepoint == "two_year_va", severity_score %in% c(73)),
    aes(label = count)
  ) +
  theme_light() + 
  labs(
    title = "VA at baseline, one, and two years by race at severity score 73",
    y = "Visual acuity"
  )
```

Patients with starting severity score of 73, split by race, with visual acuity measured over time. It appears that visual acuity changes seem to follow similar trends (i.e., net change is quite similar), but White people start with a visual acuity 1 letter higher than Black people and two letters higher than Hispanic people. 


```{r}
severity_race_va %>% 
  filter(severity_score < 80) %>% 
  ggplot(aes(x = timepoint, y = va_plot, color = severity_score, linetype = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(severity_score, race_ethnicity))) +
  theme_light() + 
  ggrepel::geom_text_repel(
    data = severity_race_va %>% filter(timepoint == "two_year_va", severity_score < 80),
    aes(label = count)
  ) +
  scale_color_viridis_c() +
  labs(
    title = "VA at baseline, one, and two years by race and severity score"
  ) 
```




```{r}
# average two year visual acuity by starting severity and race

severity_race_va %>% 
  filter(timepoint == "two_year_va") %>% 
  # filtering out severity scores larger than 79, as there are less than 50 people in each category
  filter(severity_score < 80) %>% 
  ggplot(aes(x = severity_score, y = va_plot, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = race_ethnicity)) +
  ggrepel::geom_text_repel(
    data = severity_race_va %>% filter(timepoint == "two_year_va", severity_score %in% c(35, 78)),
    aes(label = count)
  ) +
  theme_light() + 
  labs(
    title = "VA at two years by race and severity score", 
    x = "Severity score", 
    y = "Two-year visual acuity"
  )
```


This chart confirms our expectation that visual acuity two years after index date typically is lower for patients with a higher starting severity score. 

#### Timeseries VA by insurance


```{r}
# plot change in visual acuity by race at one and two yeaars

va_insurance <-
  timeseries_analysis %>% 
  group_by(insurance) %>%
  filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>% 
  summarize(
    baseline_va = mean(baseline_va_letter), 
    one_year_va = mean(one_year), 
    two_year_va = mean(two_year),
    baseline_sd = sd(baseline_va_letter),
    one_year_sd = sd(one_year),
    two_year_sd = sd(two_year),
    count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()
  
va_insurance %>% 
  ggplot(aes(x = timepoint, y = va_plot, color = insurance)) +
  geom_point() +
  geom_line(aes(group = insurance)) +
  ggrepel::geom_text_repel(
    data = va_insurance %>% filter(timepoint == "two_year_va"),
    aes(label = count)
  ) +
  theme_light() + 
  labs(
    title = "VA at baseline, one, and two years by insurance",
    y = "Visual acuity"
  )

```

```{r}

va_insurance_diff <-
  va_insurance %>% 
  spread(key = timepoint, value = va_plot) %>% 
  mutate(
    one_year_change = one_year_va - baseline_va,
    two_year_change = two_year_va - baseline_va,
    baseline = 0
  ) %>% 
  gather(key = "time_diff", value = "va_change", baseline, one_year_change, two_year_change) %>% 
  ungroup()

va_insurance_diff %>% 
  ggplot(aes(x = time_diff, y = va_change, color = insurance)) +
  geom_point() +
  geom_line(aes(group = insurance)) +
  ggrepel::geom_text_repel(
    data = va_insurance_diff %>% filter(time_diff == "two_year_change"),
    aes(label = count)
  ) +
  theme_light() + 
  labs(
    title = "VA change from baseline, one, and two years by insurance",
    y = "Visual acuity (VA) change"
  )
```


#### Percentage with 15 letter loss by race

```{r}

change_15 <- 
  timeseries_analysis %>% 
  mutate(
    one_year_change = one_year - baseline_va_letter,
    two_year_change = two_year - baseline_va_letter,
    baseline_change = 0
  ) %>% 
  mutate(
    one_year_15_pt_loss = if_else(one_year_change <= -15, 1, 0),
    two_year_15_pt_loss = if_else(two_year_change <= -15, 1, 0),
    one_year_15_pt_gain = if_else(one_year_change >= 15, 1, 0),
    two_year_15_pt_gain = if_else(two_year_change >= 15, 1, 0),
  ) %>% 
  #gather(key = "time_diff", value = "va_change", baseline_change, one_year_change, two_year_change) %>% 
  ungroup()


# z value for p = .05 
z = 1.96


loss_15_race <-
  change_15 %>% 
  group_by(race_ethnicity) %>% 
  summarize(
    one_year_loss = sum(one_year_15_pt_loss) / n(),
    two_year_loss = sum(two_year_15_pt_loss) / n(),
    one_year_sd = sqrt(one_year_loss * (1 - one_year_loss) / n()),
    two_year_sd = sqrt(two_year_loss * (1 - two_year_loss) / n())
  ) %>% 
  gather(key = "time_diff", value = "pct_15pt_loss", one_year_loss, two_year_loss) %>% 
  mutate(
    sd = if_else(str_detect(string = time_diff, pattern = "one_year"), one_year_sd, two_year_sd),
    upper_bound = pct_15pt_loss + z * sd,
    lower_bound = pct_15pt_loss - z * sd
  )
  
loss_15_race %>% 
  ggplot(aes(x = time_diff, y = pct_15pt_loss, fill = race_ethnicity)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), position = position_dodge(width = .9), width = .6) +
  theme_light() +
  labs(
    title = "Percentage of eyes that lost 15 or more pts of VA, by race"
  )


gain_15_race <-
  change_15 %>% 
  group_by(race_ethnicity) %>% 
  summarize(
    one_year_gain = sum(one_year_15_pt_gain) / n(),
    two_year_gain = sum(two_year_15_pt_gain) / n(),
    one_year_sd = sqrt(one_year_gain * (1 - one_year_gain) / n()),
    two_year_sd = sqrt(two_year_gain * (1 - two_year_gain) / n())
  ) %>% 
  gather(key = "time_diff", value = "pct_15pt_gain", one_year_gain, two_year_gain) %>% 
  mutate(
    sd = if_else(str_detect(string = time_diff, pattern = "one_year"), one_year_sd, two_year_sd),
    upper_bound = pct_15pt_gain + z * sd,
    lower_bound = pct_15pt_gain - z * sd
  )

gain_15_race %>% 
  ggplot(aes(x = time_diff, y = pct_15pt_gain, fill = race_ethnicity)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), position = position_dodge(width = .9), width = .6) +
  theme_light() +
  labs(
    title = "Percentage of eyes that gained 15 or more pts of VA, by race"
  )
```


#### Percentage with 15 letter loss by insurance

```{r}

# z value for p = .05 
z = 1.96


loss_15_insurance <-
  change_15 %>% 
  group_by(insurance) %>%
  filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>% 
  summarize(
    one_year_loss = sum(one_year_15_pt_loss) / n(),
    two_year_loss = sum(two_year_15_pt_loss) / n(),
    one_year_sd = sqrt(one_year_loss * (1 - one_year_loss) / n()),
    two_year_sd = sqrt(two_year_loss * (1 - two_year_loss) / n())
  ) %>% 
  gather(key = "time_diff", value = "pct_15pt_loss", one_year_loss, two_year_loss) %>% 
  mutate(
    sd = if_else(str_detect(string = time_diff, pattern = "one_year"), one_year_sd, two_year_sd),
    upper_bound = pct_15pt_loss + z * sd,
    lower_bound = pct_15pt_loss - z * sd
  )
  
loss_15_insurance %>% 
  ggplot(aes(x = time_diff, y = pct_15pt_loss, fill = insurance)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), position = position_dodge(width = .9), width = .6) +
  theme_light() +
  labs(
    title = "Percentage of eyes that lost 15 or more pts of VA, by insurance"
  )


gain_15_insurance <-
  change_15 %>% 
  group_by(insurance) %>% 
  filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>% 
  summarize(
    one_year_gain = sum(one_year_15_pt_gain) / n(),
    two_year_gain = sum(two_year_15_pt_gain) / n(),
    one_year_sd = sqrt(one_year_gain * (1 - one_year_gain) / n()),
    two_year_sd = sqrt(two_year_gain * (1 - two_year_gain) / n())
  ) %>% 
  gather(key = "time_diff", value = "pct_15pt_gain", one_year_gain, two_year_gain) %>% 
  mutate(
    sd = if_else(str_detect(string = time_diff, pattern = "one_year"), one_year_sd, two_year_sd),
    upper_bound = pct_15pt_gain + z * sd,
    lower_bound = pct_15pt_gain - z * sd
  )

gain_15_insurance %>% 
  ggplot(aes(x = time_diff, y = pct_15pt_gain, fill = insurance)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), position = position_dodge(width = .9), width = .6) +
  theme_light() +
  labs(
    title = "Percentage of eyes that gained 15 or more pts of VA, by insurance"
  )
```

#### Timeseries VA by insurance and race


```{r}

## NOTE - be sure to combine the two medicare datasets rather than eliminating one

# insurance_race_va %>% 
#   filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>% 
#   ggplot(aes(x = timepoint, y = va_plot, color = race_ethnicity, linetype = insurance)) +
#   geom_point() +
#   geom_line(aes(group = interaction(insurance, race_ethnicity))) +
#   ggrepel::geom_text_repel(
#     data = insurance_race_va %>% filter(timepoint == "two_year_va", insurance %in% c("Private", "Medicaid", "Medicare")), 
#     aes(label = count, hjust = -1)
#   ) +
#   theme_light() + 
#   labs(
#     title = "VA at baseline, one, and two years by race and insurance"
#   ) 
```

There are disparities in baseline VA that persist over time for Black and Hispanic patients as compared to White patients. The type of insurance appears to have strong correlation to the baseline VA, but does not fully explain the difference in race. 

```{r}
# 
# va_insurance_race_diff_tbl <-
#   insurance_race_va %>% 
#   filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>% 
#   spread(key = timepoint, value = va_plot) %>% 
#   mutate(
#     one_year_change = one_year_va - baseline_va,
#     two_year_change = two_year_va - baseline_va,
#     baseline = 0
#   )
# 
# va_insurance_race_diff <-
#   insurance_race_va %>% 
#   filter(!insurance %in% c("Unknown/Missing", "Military", "Govt")) %>% 
#   spread(key = timepoint, value = va_plot) %>% 
#   mutate(
#     one_year_change = one_year_va - baseline_va,
#     two_year_change = two_year_va - baseline_va,
#     baseline = 0
#   ) %>% 
#   gather(key = "time_diff", value = "va_change", baseline, one_year_change, two_year_change) %>% 
#   ungroup()
# 
# va_insurance_race_diff_ci <-
#   timeseries_analysis %>% 
#   mutate(
#     one_year_change = one_year - baseline_va_letter,
#     two_year_change = two_year - baseline_va_letter,
#     baseline = 0
#   ) %>% 
#   group_by(race_ethnicity, insurance) %>% 
#   summarize(
#     mean_one_year_change = mean(one_year_change),
#     mean_two_year_change = mean(two_year_change),
#     sd_one_year_change = sd(one_year_change),
#     sd_two_year_change = sd(two_year_change)
#   )
# 
# va_insurance_race_diff %>% 
#   ggplot(aes(x = time_diff, y = va_change, color = race_ethnicity, linetype = insurance)) +
#   geom_point() +
#   geom_line(aes(group = interaction(insurance, race_ethnicity))) +
#   ggrepel::geom_text_repel(
#     data = va_insurance_race_diff %>% filter(time_diff == "two_year_change", insurance %in% c("Private", "Medicaid", "Medicare")), 
#     aes(label = count, hjust = -2)
#   ) +
#   facet_grid(insurance~.) +
#   theme_light() + 
#   labs(
#     title = "VA change from baseline, one, and two years by insurance and race",
#     y = "Visual acuity (VA) change"
#   )


```



#### Timeseries VA by smoking and race

```{r}
smoking_race_va <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, smoke_status) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va)

```

```{r}

## NOTE - be sure to combine the two medicare datasets rather than eliminating one

smoking_race_va %>% 
  #filter(!smoke_status %in% c("Unknown / Unclassified")) %>% 
  ggplot(aes(x = timepoint, y = va_plot, color = race_ethnicity, linetype = smoke_status)) +
  geom_point() +
  geom_line(aes(group = interaction(smoke_status, race_ethnicity))) +
  ggrepel::geom_text_repel(
    data = smoking_race_va %>% filter(timepoint == "two_year_va"), #!smoke_status %in% c("Unknown / Unclassified") 
    aes(label = count, hjust = -1)
  ) +
  theme_light() + 
  labs(
    title = "VA at baseline, one, and two years by race and smoking status"
  ) 
```

#### Timeseries VA by gender and race

```{r}
gender_race_va <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, gender) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va)

```

```{r}
gender_race_va %>% 
  filter(gender %in% c("Male", "Female")) %>% 
  ggplot(aes(x = timepoint, y = va_plot, color = race_ethnicity, linetype = gender)) +
  geom_point() +
  geom_line(aes(group = interaction(gender, race_ethnicity))) +
  ggrepel::geom_text_repel(
    data = gender_race_va %>% filter(timepoint == "two_year_va", gender %in% c("Male", "Female")), 
    aes(label = count, hjust = -1)
  ) +
  theme_light() + 
  labs(
    title = "VA at baseline, one, and two years by race and gender"
  ) 
```

Female patients appear to have lower baseline VA as compared to their male counterparts of the same race. The raw gain or loss in VA over time appears to be similar between genders within racial groups.


#### Race VA timeseries

```{r}
va_race_tbl <-
  timeseries_analysis %>% 
  group_by(race_ethnicity) %>% 
  summarize(
    mean_baseline_va = mean(baseline_va_letter),
    mean_one_year = mean(one_year),
    mean_two_year = mean(two_year),
    count = n()
  )
```


```{r}

race_va <-
  timeseries_analysis %>% 
  # round baseline data to nearest multiple of 5
  mutate(
    baseline_va_r = plyr::round_any(baseline_va_letter, 5)
  ) %>% 
  # group patients by race and same baseline va data
  group_by(baseline_va_r, race_ethnicity) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()

```

```{r}
race_va %>%
  filter(count > 25) %>% 
  ggplot(aes(x = timepoint, y = va_plot, color = baseline_va_r, linetype = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(baseline_va_r, race_ethnicity))) +
  theme_light() + 
  scale_color_viridis_c() +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at the same baseline visual acuity"
  ) 
```

This graph is hard to interpret, but is showing the differential progression of va for people of different races that start at the same level of baseline visual acuity. The next graphs dive into this daata more closely.

This graph takes on an interesting shape. It indicates that people that start with higher visual acuity are likely to experience a decline in visual acuity, aand those starting at the lowest end of the spectrum of visuala acuity are likely to make gains.

```{r}
race_va %>%
  filter(count > 25) %>% 
  filter(baseline_va_r %in% c(65, 70, 75)) %>% 
  mutate(baseline_va_r = as.factor(baseline_va_r)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = baseline_va_r, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(baseline_va_r, race_ethnicity))) +
  theme_light() + 
  ggrepel::geom_text_repel(
    data = race_va %>% filter(timepoint %in% c("two_year_va"), baseline_va_r %in% c(65, 70, 75)) %>% mutate(baseline_va_r = as.factor(baseline_va_r)), 
    aes(label = count, hjust = -1)
  ) +
  scale_y_continuous(breaks = seq(64, 78, 1)) +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at ~65, ~70, ~75 baseline va",
    y = "visual acuity (letter)"
  ) 
```

We observe that Hispanic patients experience starting with ~65 visual acuity, they experiences much less gain in vision over two years as compared to their White and Black counterprats. White people appear to gain two letters, while Hispanic people gain less than hallf a letter.

Exemplifying the trends from above, we also se that those who start with VA at ~75 all experience substantial decline, with those at other starting points experiencing less severe declines or even making gains.



```{r}
race_va %>%
  filter(count > 25) %>% 
  filter(baseline_va_r %in% c(50, 55, 60)) %>% 
  mutate(baseline_va_r = as.factor(baseline_va_r)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = baseline_va_r, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(baseline_va_r, race_ethnicity))) +
  theme_light() + 
  ggrepel::geom_text_repel(
    data = race_va %>% filter(timepoint %in% c("two_year_va"), baseline_va_r %in% c(50, 55, 60)) %>% mutate(baseline_va_r = as.factor(baseline_va_r)), 
    aes(label = count, hjust = -1)
  ) +
  scale_y_continuous(breaks = seq(40, 65, 1)) +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at ~50, ~55, ~60 baseline va",
    y = "visual acuity (letter)"
  ) 
```


At this level, the dispairities in improvements are much more stark. Whie people with baseline va of 50 are 3 additional points of VA gain as compared to their Black and Hispanic counterparts. Similar, though not as stark disparities are present for people with starting VA at 55 and 60.


Further analysis is required to control for the specific condition associated that the patient is diagnosed with. 

```{r}
race_va %>%
  filter(count > 25) %>% 
  filter(baseline_va_r %in% c(35, 40, 45)) %>% 
  mutate(baseline_va_r = as.factor(baseline_va_r)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = baseline_va_r, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(baseline_va_r, race_ethnicity))) +
  theme_light() + 
  ggrepel::geom_text_repel(
    data = race_va %>% filter(timepoint %in% c("two_year_va"), baseline_va_r %in% c(35, 40, 45)) %>% mutate(baseline_va_r = as.factor(baseline_va_r)), 
    aes(label = count, hjust = -1)
  ) +
  scale_y_continuous(breaks = seq(30, 60, 1)) +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at ~35, ~40, ~45 baseline va",
    y = "visual acuity (letter)"
  ) 
```


#### Race VA timeseries bucketed by 10


```{r}

race_va_10 <-
  timeseries_analysis %>% 
  # group patients by race and same baseline va data
  group_by(baseline_va_r10, race_ethnicity) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()


## Same as above - split by ten 
race_va_10 %>%
  filter(count > 25) %>% 
  filter(baseline_va_r10 %in% c(50, 60, 70)) %>% 
  mutate(baseline_va_r10 = as.factor(baseline_va_r10)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = baseline_va_r10, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(baseline_va_r10, race_ethnicity))) +
  theme_light() + 
  ggrepel::geom_text_repel(
    data = race_va_10 %>% filter(timepoint %in% c("two_year_va"), baseline_va_r10 %in% c(50, 60, 70)) %>% mutate(baseline_va_r10 = as.factor(baseline_va_r10)), 
    aes(label = count, hjust = -1)
  ) +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at ~50, ~60, ~70 baseline va",
    y = "visual acuity (letter)"
  ) 
```

```{r}

severity_race_va_10 <-
  timeseries_analysis %>% 
  # group patients by race, severity, and same baseline va data
  group_by(baseline_va_r10, severity_score, race_ethnicity) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()


severity_race_va_10_sd <-
  timeseries_analysis %>% 
  # group patients by race, severity, and same baseline va data
  group_by(baseline_va_r10, severity_score, race_ethnicity) %>% 
  summarize(baseline_va = sd(baseline_va_letter), one_year_va = sd(one_year), two_year_va = sd(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot_sd", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()

severity_race_va_10 %>% 
  arrange(desc(count))

```

```{r}
severity_race_va_10 %>%
  filter(
    baseline_va_r10 %in% c(80),
    severity_score %in% c(73)
  ) %>% 
  left_join(severity_race_va_10_sd %>% select(-count), by = c("baseline_va_r10", "severity_score", "race_ethnicity", "timepoint")) %>% 
  mutate(severity_score = as.factor(severity_score)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = severity_score, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(severity_score, race_ethnicity))) +
  theme_light() + 
  ggrepel::geom_text_repel(
    data = severity_race_va_10 %>% filter(timepoint %in% c("two_year_va"), baseline_va_r10 %in% c(80), severity_score %in% c(73)) %>% mutate(severity_score = as.factor(severity_score)), 
    aes(label = count, hjust = -1)
  ) +
  geom_errorbar(aes(ymin = va_plot - va_plot_sd, ymax = va_plot + va_plot_sd)) +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at ~80 baseline VA",
    subtitle = "Grouped by starting severity score and race",
    y = "visual acuity (letter)"
  ) 
```




#### Severity race VA timeseries

```{r}

severity_race_va <-
  timeseries_analysis %>% 
  # round baseline data to nearest multiple of 5
  mutate(
    baseline_va_r = plyr::round_any(baseline_va_letter, 5)
  ) %>% 
  # group patients by race, severity, and same baseline va data
  group_by(baseline_va_r, severity_score, race_ethnicity) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()

```

```{r}
severity_race_va %>%
  filter(
    baseline_va_r %in% c(50),
    severity_score %in% c(73, 78)
  ) %>% 
  mutate(severity_score = as.factor(severity_score)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = severity_score, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(severity_score, race_ethnicity))) +
  theme_light() + 
  ggrepel::geom_text_repel(
    data = severity_race_va %>% filter(timepoint %in% c("two_year_va"), baseline_va_r %in% c(50), severity_score %in% c(73, 78)) %>% mutate(severity_score = as.factor(severity_score)), 
    aes(label = count, hjust = -1)
  ) +
  scale_y_continuous(breaks = seq(40, 65, 1)) +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at ~50 baseline VA",
    subtitle = "Grouped by starting severity score and race",
    y = "visual acuity (letter)"
  ) 
```

For patients that start with a baseline visual acuity close to 50, the change in visual acuity over one to two years is associated with race. White patients with starting severity at 73 experience ~6pt gain in visual acuity, as compared to ~5pt gain for Black and Hispanic counterparts. After two years, white patients maintain that gain in visual acuity; however, Black and Hispanic patients experiences a moderate drop in visual acuity by .5pts (Hispanics) and 1.5pts (Black).

For those patients with starting severity at 78, gains in visual acuity the first year are again similar across races, ~3 to 3.5 points. However, in year two, white patients experience another substantial gain in visual acuity (~3pts), while Hispanic paitents experience no gain in visual acuity and Black patients experience a small decline in visual acuity (~.5 pts). 

This analysis suggests that there may be factors between the first and second year of treatment that are creating disparate treatment outcomes for Black and Hispanic patients as compared to white patients. 


```{r}
severity_race_va %>%
  filter(
    baseline_va_r %in% c(35),
    severity_score %in% c(73, 78)
  ) %>% 
  mutate(severity_score = as.factor(severity_score)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = severity_score, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(severity_score, race_ethnicity))) +
  theme_light() + 
  theme(panel.grid.minor = element_blank()) +
  ggrepel::geom_text_repel(
    data = severity_race_va %>% filter(timepoint %in% c("two_year_va"), baseline_va_r %in% c(35), severity_score %in% c(73, 78)) %>% mutate(severity_score = as.factor(severity_score)), 
    aes(label = count, hjust = -1)
  ) +
  scale_y_continuous(breaks = seq(33, 55, 1)) +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at ~35 baseline VA",
    subtitle = "Grouped by starting severity score and race",
    y = "visual acuity (letter)"
  ) 
```

The trend identified for patients with a baseline VA of 50 (the prior chart) partially holds for those patients with starting VA of 35. We see that, for patients with starting severity of 73, gains in visual acuity after one year are consistent by race. However, aafter two yers, Black patients lost abaout 3pts of visual acutiy, while white patients gained ~1pt in visual acuity and Hispanic patients did not experience any change.

This trend does not hold for patients with starting severity of 78. We see that in year one, white patients make substantially higher gains (16pts) in VA as compared to Black patients (14 pts) and Hipanic patients (11 pts). After two years, however, white patients experience a drop in visual acuity (1 pts) while Black patients experience another gain of three points. 



```{r}
severity_race_va %>%
  filter(
    baseline_va_r %in% c(70),
    severity_score %in% c(73, 78)
  ) %>% 
  mutate(severity_score = as.factor(severity_score)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = severity_score, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(severity_score, race_ethnicity))) +
  theme_light() + 
  theme(panel.grid.minor = element_blank()) +
  ggrepel::geom_text_repel(
    data = severity_race_va %>% filter(timepoint %in% c("two_year_va"), baseline_va_r %in% c(70), severity_score %in% c(73, 78)) %>% mutate(severity_score = as.factor(severity_score)), 
    aes(label = count, hjust = -1)
  ) +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at ~70 baseline VA",
    subtitle = "Grouped by starting severity score and race",
    y = "visual acuity (letter)"
  ) 
```



#### Severity insurance race VA timeseries

```{r}

severity_ins_race_va <-
  timeseries_analysis %>% 
  # round baseline data to nearest multiple of 5
  mutate(
    baseline_va_r = plyr::round_any(baseline_va_letter, 5)
  ) %>% 
  # group patients by race, severity, and same baseline va data
  group_by(baseline_va_r, severity_score, insurance, race_ethnicity) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()

```

```{r}
severity_ins_race_va %>%
  filter(
    baseline_va_r %in% c(50),
    severity_score %in% c(78),
    insurance %in% c("Medicaid", "Medicare", "Private")
  ) %>% 
  mutate(severity_score = as.factor(severity_score)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = insurance, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(insurance, race_ethnicity))) +
  theme_light() + 
  ggrepel::geom_text_repel(
    data = severity_ins_race_va %>% filter(timepoint %in% c("two_year_va"), baseline_va_r %in% c(50), severity_score %in% c(78), insurance %in% c("Medicaid", "Medicare", "Private")) %>% mutate(severity_score = as.factor(severity_score)), 
    aes(label = count, hjust = -1)
  ) +
  labs(
    title = "VA at baseline, one, and two years by race for people starting at ~50 baseline VA",
    subtitle = "Grouped by insuraance and race; starting severity = 78 ",
    y = "visual acuity (letter)"
  ) 
```







## Injections by VA analysis

```{r, eval = FALSE}
## calculate change in va over time for people in each severity group/race
inj_race_va <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, inj_year_1) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>%
  filter(inj_year_1 < 13) %>% 
  mutate(inj_year_1 = as.factor(inj_year_1)) %>% 
  ungroup()


str(inj_race_va)
```


```{r, eval = FALSE}

inj_race_va %>%
  filter(inj_year_1 %in% c(1, 8)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, color = race_ethnicity, linetype = inj_year_1)) +
  geom_point() +
  geom_line(aes(group = interaction(inj_year_1, race_ethnicity))) +
  ggrepel::geom_text_repel(
    data = inj_race_va %>% filter(timepoint == "two_year_va", inj_year_1 %in% c(1, 8)),
    aes(label = count, hjust = -1)
  ) +
  theme_light() + 
  labs(
    title = "VA at baseline, one, and two years by race and number of injections"
  ) 
```


```{r, eval = FALSE}

severity_race_inj_va_quart <-
  timeseries_analysis %>% 
  # group patients by race, severity, and same baseline va data
  group_by(baseline_va_quart, severity_score, race_ethnicity, inj_year_1) %>% 
  summarize(baseline_va = mean(baseline_va_letter), one_year_va = mean(one_year), two_year_va = mean(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()


severity_race_inj_va_quart_sd <-
  timeseries_analysis %>% 
  # group patients by race, severity, and same baseline va data
  group_by(baseline_va_quart, severity_score, race_ethnicity, inj_year_1) %>% 
  summarize(baseline_va = sd(baseline_va_letter), one_year_va = sd(one_year), two_year_va = sd(two_year), count = n()) %>% 
  gather(key = "timepoint", value = "va_plot_sd", baseline_va, one_year_va, two_year_va) %>% 
  ungroup()


severity_race_inj_va_quart %>% 
  arrange(desc(count))

```

```{r, eval = FALSE}
severity_race_inj_va_quart %>%
  filter(
    baseline_va_quart %in% c(4),
    severity_score %in% c(73),
    inj_year_1 %in% c(1)
  ) %>% 
  mutate(severity_score = as.factor(severity_score)) %>% 
  ggplot(aes(x = timepoint, y = va_plot, linetype = severity_score, color = race_ethnicity)) +
  geom_point() +
  geom_line(aes(group = interaction(severity_score, race_ethnicity))) +
  theme_light() + 
  ggrepel::geom_text_repel(
    data = 
      severity_race_inj_va_quart %>% 
      filter(
        timepoint %in% c("two_year_va"), 
        baseline_va_quart %in% c(4), 
        severity_score %in% c(73), 
        inj_year_1 %in% c(1)) %>% 
      mutate(severity_score = as.factor(severity_score)), 
    aes(label = count, hjust = -1)
  ) +
  labs(
    title = "VA at baseline, one, and two years by race for people at ~4th VA quart, 73 severity",
    subtitle = "Grouped by starting severity score, race, and number of injections = 1",
    y = "visual acuity (letter)"
  ) 
```

Hispanic patients with the same starting severity and baseline VA quartile that receive the same number of injections in the first year (1 injection) do substantially worse than their white counterparts



## Baseline analysis
#### Severity by race

```{r}

# Calculate severity by race
severity_race <-
  timeseries_analysis %>% 
  group_by(race_ethnicity) %>% 
  summarize(average = mean(severity_score), count = n()) %>% 
  arrange(desc(average))

# Output severity by race to a cleaner table
severity_race %>% kable()
```

Hispanic people have the highest severity of all racial groups, nearaly 5 points higher than White people. Black, Asian, and Caucasian people have similar severity at timme of diagnosis.

#### Severity by race and region

```{r}

# Calculate severity by race and region
severity_race_reg <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, region) %>% 
  summarize(average = mean(severity_score), count = n()) %>% 
  arrange(desc(average))

# Print severity by race and region table
severity_race_reg

```


```{r}

# graph severity by race and region

severity_race_reg %>% 
  filter(!race_ethnicity == "Unknown") %>% 
  ggplot(aes(x = region, y = average)) + 
  geom_col(aes(fill  = race_ethnicity), position = "dodge") +
  scale_fill_viridis_d() + 
  theme_light() +
  labs(
    title = "Severity by race and region"
  )
```

While there is some regional variation, Hispanic people are most likely among all racial groups to have higher severity at time of diagnosis as compared to other racial groups.

#### Severity by race and age

```{r}

#calculate severity by race and age

severity_race_age <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, age_group) %>% 
  summarize(avg_severity = mean(severity_score), count = n()) %>% 
  arrange(desc(avg_severity))

severity_race_age %>% head(10) %>% kable()

```

```{r}
# graph severity by race and age group

severity_race_age %>%
  filter(race_ethnicity %in% c("Caucasian", "Black or African American", "Hispanic", "Asian")) %>% 
  ggplot(aes(age_group, avg_severity)) +
  geom_line(aes(color = race_ethnicity)) +
  geom_point(aes(color = race_ethnicity)) +
  theme_light() +
  labs(
    title = "Severity by race and age group"
  )
```

[All data]
Analysis

[New classification]
We see that at all age levels between 45 to 80, Hispanic people have higher severity at time of diagnosis. 
Asian, Black, and Caucasian people appear to have similar severity at time of diagnosis, though a slight gap emerges at around age 75. 


```{r}
# close-up look at severity by race and age
severity_race_age %>%
  filter(race_ethnicity %in% c("Caucasian", "Black or African American", "Hispanic")) %>% 
  filter(age_group > 65, age_group < 95) %>% 
  ggplot(aes(age_group, avg_severity)) +
  geom_line(aes(color = race_ethnicity)) +
  geom_point(aes(color = race_ethnicity)) +
  theme_light() +
  theme(
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(breaks = c(48, 50, 52, 54, 56)) + 
  labs(
    title = "Severity by race and age group, between 70 to 90 years old", 
    y = "Average severity score at time of diagnosis", 
    x = "Age group (nearest multiple of 5)"
  )
```

[New classification]
Average severity at time of diagnosis is ~1.5-2 pts higher for Black people vs. Caucasian people, and ~3-4 pts higher for Hispanic poeple as compared to Caucasian people

#### Severity by region

```{r}

# severity by region

severity_reg <-
  timeseries_analysis %>% 
  group_by(region) %>% 
  summarize(average = mean(severity_score), n = n()) %>% 
  arrange(desc(average))

severity_reg %>% kable()

```

Severity in the West does appear to be a few points higher than in other regions - perhaps due to a higher number of Hispanic people being from the West.


#### Severity by race and insurance

```{r}
# severity by race and insurace

severity_race_ins <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, insurance) %>% 
  summarize(average = mean(severity_score), count = n()) %>% 
  arrange(desc(average))

severity_race_ins
```

```{r}

# graph severity by race and insurance 

severity_race_ins %>% 
  filter(!race_ethnicity %in% c("Unknown", "Other")) %>% 
  ggplot(aes(x = insurance, y = average)) + 
  geom_col(aes(fill  = race_ethnicity), position = "dodge") +
  theme_light() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_viridis_d() + 
  labs(
    title = "Severity by race and insurance",
    y = "Average severity at time of diagnosis",
    x = "Type of insurance"
  )
```

We see large Hispanic/white gaps for those on Govt,Medicare FFS, Private or with Unknown/Missing insurance information. Smaller Hispanic/white gap for those on Medicaid or with Military insurance.


#### Severity by race and sex

```{r}

# severity by race and sex

severity_race_sex <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, gender) %>% 
  summarize(average = mean(severity_score), count = n()) %>% 
  arrange(desc(average))

severity_race_sex %>% head(5) %>% kable()
```


```{r}

# graphing severity by race and sex
severity_race_sex %>% 
  filter(!gender %in% c("Unknown")) %>% 
  ggplot(aes(x = race_ethnicity, y = average)) + 
  geom_point(aes(color = gender), size = 2) +
  theme_light() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_viridis_d() + 
  labs(
    title = "Severity by race and sex"
  )
```

Male patients have higher severity at the time of diagnosis than female patients among all racial groups.


#### Severity by type of treatment

```{r}

# severity by race and type of treatment

severity_race_first_treatment <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, proc_group_28) %>% 
  summarize(average = mean(severity_score), count = n()) %>% 
  arrange(desc(average))

# severity_race_first_treatment
```

```{r}

# graphing severity by race and type of treatment
severity_race_first_treatment %>% 
  ggplot(aes(x = race_ethnicity, y = average)) + 
  geom_point(aes(color = proc_group_28), size = 2) +
  theme_light() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_viridis_d() + 
  labs(
    title = "Severity by race and first treatment type (28 days)"
  )
```

Physicians are treating people with higher severity within their racial group with antivegf or combo drugs for their first treatment, as compared to laser or steroid. 


```{r}
# Likelihood of first treatment type by severity and race
treatment_sev_race <-
  timeseries_analysis %>% 
  count(severity_score, race_ethnicity, proc_group_28) %>% 
  group_by(severity_score, race_ethnicity) %>% 
  mutate(count_severity_race = sum(n)) %>% 
  mutate(prop = n / count_severity_race) %>% 
  ungroup() %>% 
  select(severity_score, race_ethnicity, proc_group_28, prop, count = n)

treatment_sev_race %>% 
  arrange(desc(severity_score, race_ethnicity))
```

```{r}

# Graphing likelihood of first treatment type by severity and race
treatment_sev_race %>% 
  filter(!race_ethnicity %in% c("Unknown", "Asian")) %>% 
  filter(severity_score > 50) %>% 
  ggplot(aes(x = severity_score, y = prop, color = race_ethnicity, shape = proc_group_28)) +
  geom_point() + 
  #facet_wrap("proc_group_28") +
  theme_bw() +
  #scale_y_continuous(labels = scales::percent, breaks = c(.70, .75, .8, .85, .9), limits = c(.7, .9)) +
  labs(
    title = "Likelihood of treatment at each severity score by race",
    y = "Percentage of people receiving this treatment"
  )

```
[New classification]
At this level, it is difficult to see how race may be impacting treatment assigment. In the next graph, we look more closely at the likelihood of receiving an anti-vegf treatment, based on race. 

```{r}

# Graphing likelihood of antivegf as first treatment by severity and race
treatment_sev_race %>% 
  filter(!race_ethnicity %in% c("Unknown", "Asian")) %>% 
  filter(severity_score > 50) %>% 
  filter(proc_group_28 == "antivegf") %>% 
  ggplot(aes(x = severity_score, y = prop, color = race_ethnicity)) +
  geom_point() + 
  #facet_wrap("proc_group_28") +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of antivegf at each severity score by race",
    y = "Percentage of people receiving antivegf"
  )

```

[New classification]
Black people appear to be 2-5% points less likely to receive antivegf treatment as compared to Caucasian people at several severity scores (53, 58, 73, 78)


We can extend this analysis by looking at which antivegf drugs patients of different races receive, at different severity levels.

#### Likelihood of drug received by severity score and race (given that patient recieved anti-vegf treatment)


```{r}

# Calculate likelihood of drug by severity score and race (given that patient received antivegf treatment)
drug_sev_race <-
  timeseries_analysis %>% 
  filter(proc_group_365 == "antivegf") %>% 
  count(severity_score, race_ethnicity, vegf_group_365) %>% 
  group_by(severity_score, race_ethnicity) %>% 
  mutate(count_severity_race = sum(n)) %>% 
  mutate(prop = n / count_severity_race) %>% 
  ungroup() %>% 
  select(severity_score, race_ethnicity, vegf_group_365, prop, count = n)

drug_sev_race %>% 
  arrange(desc(severity_score, race_ethnicity))

```

```{r}

# Graphing likelihood of first drug type by severity and race
drug_sev_race %>% 
  filter(!race_ethnicity %in% c("Unknown", "Asian", "Other")) %>% 
  ggplot(aes(x = severity_score, y = prop, color = race_ethnicity)) +
  geom_point() + 
  theme_bw() +
  facet_grid(~vegf_group_365) +
  #scale_y_continuous(labels = scales::percent, breaks = c(.70, .75, .8, .85, .9), limits = c(.7, .9)) +
  labs(
    title = "Likelihood of drug at each severity score by race",
    y = "Percentage of people receiving this drug"
  )

```

```{r}

# Graphing likelihood of Eylea as first treatment by severity and race
drug_sev_race %>% 
  filter(!race_ethnicity %in% c("Unknown", "Other")) %>% 
  filter(severity_score > 50) %>% 
  filter(vegf_group_365 == "Eylea") %>% 
  ggplot(aes(x = severity_score, y = prop, color = race_ethnicity)) +
  geom_point() + 
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Eylea at higher severity score by race",
    y = "Percentage of people receiving Eylea"
  )

```

Caucasian people are 5-10% more likely to receive Eylea treatment (first) as compared to Hispanic people at the same level of severity, and 2-5% more likely to receive Eylea treatment (first) as compared to Black people at the same level of severity. 


```{r}

# Graphing likelihood of Bevacizumab as first treatment by severity and race
drug_sev_race %>% 
  filter(!race_ethnicity %in% c("Unknown", "Other")) %>% 
  filter(severity_score < 50) %>% 
  filter(vegf_group_365 == "Avastin") %>% 
  ggplot(aes(x = severity_score, y = prop, color = race_ethnicity)) +
  geom_point() + 
  geom_line() +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Bevacizumab at higher severity score by race",
    y = "Percentage of people receiving Bevacizumab"
  )

```

At lower levels of severity, Hispanic people are far more likely (15-25%) to receive Bevacizumab as compared to their Caucasian counterparts with the same level of severity. Black and Asian people seem somewhat more likely to receive Bevacizumab, but there is some variation. 


#### Likelihood of drug received by severity score CATEGORY and race (given that patient recieved anti-vegf treatment)


```{r}

# Calculate likelihood of drug by severity score and race (given that patient received antivegf treatment)
drug_sev_cat_race <-
  timeseries_analysis %>% 
  filter(proc_group_365 == "antivegf") %>% 
  count(vision_category, race_ethnicity, vegf_group_365) %>% 
  group_by(vision_category, race_ethnicity) %>% 
  mutate(count_severity_race = sum(n)) %>% 
  mutate(prop = n / count_severity_race) %>% 
  ungroup() %>% 
  select(vision_category, race_ethnicity, vegf_group_365, prop, count = n)

drug_sev_cat_race %>% 
  arrange(desc(vision_category, race_ethnicity))

```

```{r}

# Graphing likelihood of first drug type by severity and race
drug_sev_cat_race %>% 
  filter(!race_ethnicity %in% c("Unknown", "Asian", "Other"), !is.na(vision_category), !vision_category %in% c("NA")) %>% 
  mutate(prop_round = round(prop, 2)) %>% 
  ggplot(aes(x = vision_category, y = prop, fill = race_ethnicity)) +
  geom_col(position = "dodge") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_grid(vegf_group_365~.) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_text(aes(label = prop_round), vjust = -0.5, position = position_dodge(width = .9)) +
  #scale_x_continuous()
  #scale_y_continuous(labels = scales::percent, breaks = c(.70, .75, .8, .85, .9), limits = c(.7, .9)) +
  labs(
    title = "Likelihood of drug at each severity score by race",
    y = "Percentage of people receiving this drug"
  )

```

```{r}

# Graphing likelihood of Eylea as first treatment by severity and race
drug_sev_cat_race %>% 
  filter(!is.na(vision_category)) %>% 
  filter(vegf_group_365 == "Eylea") %>% 
  ggplot(aes(x = vision_category, y = prop, color = race_ethnicity)) +
  geom_point() + 
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Eylea at higher vision_category by race",
    y = "Percentage of people receiving Eylea"
  )

```


```{r}

# Graphing likelihood of Bevacizumab as first treatment by severity and race
drug_sev_cat_race %>% 
  filter(!is.na(vision_category)) %>% 
  filter(vegf_group_365 == "Avastin") %>% 
  ggplot(aes(x = vision_category, y = prop, color = race_ethnicity)) +
  geom_point() + 
  geom_line() +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Bevacizumab at higher vision_category by race",
    y = "Percentage of people receiving Bevacizumab"
  )

```



#### Likelihood of drug received by baseline VA and race (given that patient recieved anti-vegf treatment)

```{r}

# Calculate likelihood of drug by severity score and race (given that patient received antivegf treatment)
drug_va_race <-
  timeseries_analysis %>% 
  filter(proc_group_365 == "antivegf") %>% 
  count(baseline_va_r10, race_ethnicity, vegf_group_365) %>% 
  group_by(baseline_va_r10, race_ethnicity) %>% 
  mutate(count_va_race = sum(n)) %>% 
  mutate(prop = n / count_va_race) %>% 
  ungroup() %>% 
  select(baseline_va_r10, race_ethnicity, vegf_group_365, prop, count = n)

drug_va_race %>% 
  arrange(desc(baseline_va_r10, race_ethnicity))

```

```{r}

# Graphing likelihood of first drug type by severity and race
drug_va_race %>% 
  filter(baseline_va_r10 <= 80) %>% 
  ggplot(aes(x = baseline_va_r10, y = prop, color = race_ethnicity)) +
  geom_point() + 
  geom_line() +
  theme_bw() +
  facet_grid(vegf_group_365~.) +
  #scale_y_continuous(labels = scales::percent, breaks = c(.70, .75, .8, .85, .9), limits = c(.7, .9)) +
  labs(
    title = "Likelihood of drug at each baseline va (rounded) by race",
    y = "Percentage of people receiving this drug"
  )

```



#### Likelihood of drug received by severity score CATEGORY and insurance (given that patient recieved anti-vegf treatment)


```{r}

# Calculate likelihood of drug by severity score and insurance (given that patient received antivegf treatment)
drug_sev_cat_insurance <-
  timeseries_analysis %>%
  filter(proc_group_365 == "antivegf", insurance %in% c("Private", "Medicare", "Medicaid")) %>% 
  count(vision_category, insurance, vegf_group_365) %>% 
  group_by(vision_category, insurance) %>% 
  mutate(count_severity_race = sum(n)) %>% 
  mutate(prop = n / count_severity_race) %>% 
  ungroup() %>% 
  select(vision_category, insurance, vegf_group_365, prop, count = n)

drug_sev_cat_insurance %>% 
  arrange(desc(vision_category, insurance))

```

```{r}

# Graphing likelihood of first drug type by severity and insurance
drug_sev_cat_insurance %>% 
  filter(!is.na(vision_category), !vision_category %in% c("NA")) %>% 
  mutate(prop_round = round(prop, 2)) %>% 
  ggplot(aes(x = vision_category, y = prop, fill = insurance)) +
  geom_col(position = "dodge") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_grid(vegf_group_365~.) +
  scale_y_continuous(limits = c(0, 1)) +
  #geom_text(aes(label = prop_round), vjust = -0.5, position = position_dodge(width = .9)) +
  #scale_x_continuous()
  #scale_y_continuous(labels = scales::percent, breaks = c(.70, .75, .8, .85, .9), limits = c(.7, .9)) +
  labs(
    title = "Likelihood of drug at each severity score by insurance",
    y = "Percentage of people receiving this drug"
  )

```

```{r}

# Graphing likelihood of Eylea as first treatment by severity and insurance
drug_sev_cat_insurance %>% 
  filter(!is.na(vision_category)) %>% 
  filter(vegf_group_365 == "Eylea") %>% 
  ggplot(aes(x = vision_category, y = prop, color = insurance)) +
  geom_point() + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Eylea at higher vision_category by race",
    y = "Percentage of people receiving Eylea"
  )

```


```{r}

# Graphing likelihood of Bevacizumab as first treatment by severity and insurance
drug_sev_cat_insurance %>% 
  filter(!is.na(vision_category)) %>% 
  filter(vegf_group_365 == "Avastin") %>% 
  ggplot(aes(x = vision_category, y = prop, color = insurance)) +
  geom_point() + 
  geom_line() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Bevacizumab at higher vision_category by race",
    y = "Percentage of people receiving Bevacizumab"
  )

```

#### Drug received by race and insurance

Part of the reason there may be variation by race in which treatment drug is received is due to insurance. Medical providers may only choose drugs covered by the patients insurance (e.g., Medicare does not cover Eylea). 

```{r}
# Calculate likelihood of drug by severity score and race (given that patient received antivegf treatment)
drug_sev_race_ins <-
  timeseries_analysis %>% 
  filter(proc_group_28 == "antivegf") %>% 
  count(severity_score, race_ethnicity, vegf_group_28, insurance) %>% 
  group_by(severity_score, race_ethnicity, insurance) %>% 
  mutate(count_severity_race = sum(n)) %>% 
  mutate(prop = n / count_severity_race) %>% 
  ungroup() %>% 
  select(severity_score, race_ethnicity, insurance, vegf_group_28, prop, count = n)

drug_sev_race_ins %>% 
  arrange(desc(severity_score, race_ethnicity, insurance))

```

```{r}
# Graphing likelihood of Eylea as first treatment by severity, race, and Medicare FFS  insurance
drug_sev_race_ins %>% 
  filter(!race_ethnicity %in% c("Unknown", "Other", "Asian")) %>% 
  filter(severity_score > 50) %>% 
  filter(vegf_group_28 == "Eylea") %>% 
  filter(insurance == "Medicare") %>% 
  ggplot(aes(x = severity_score, y = prop, color = race_ethnicity)) +
  geom_point() + 
  geom_line() +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Eylea at higher severity score by race, with Medicare insurance",
    y = "Percentage of people receiving Eylea"
  )

```


```{r}
# Graphing likelihood of Eylea as first treatment by severity, race, and Medicaid insurance
drug_sev_race_ins %>% 
  # Asian filtered due to too small data size
  filter(!race_ethnicity %in% c("Unknown", "Other", "Asian")) %>% 
  filter(severity_score > 50) %>% 
  filter(vegf_group_28 == "Eylea") %>% 
  filter(insurance == "Medicaid") %>% 
  ggplot(aes(x = severity_score, y = prop, color = race_ethnicity)) +
  geom_point() + 
  geom_line() +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Eylea at higher severity score by race, with Medicaid insurance",
    y = "Percentage of people receiving Eylea"
  )

```

```{r}
# Graphing likelihood of Eylea as first treatment by severity, race, and private insurance
drug_sev_race_ins %>% 
  # Asian filtered due to too small data size
  filter(!race_ethnicity %in% c("Unknown", "Other", "Asian")) %>% 
  filter(severity_score > 50) %>% 
  filter(vegf_group_28 == "Eylea") %>% 
  filter(insurance == "Private") %>% 
  ggplot(aes(x = severity_score, y = prop, color = race_ethnicity)) +
  geom_point() + 
  geom_line() +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Eylea at higher severity score by race, with private insurance",
    y = "Percentage of people receiving Eylea"
  )

```

```{r}
# Calculate likelihood of drug by severity score, race, baseline va (given that patient received antivegf treatment)
drug_sev_race_va <-
  timeseries_analysis %>% 
  filter(proc_group_28 == "antivegf") %>% 
  count(severity_score, race_ethnicity, vegf_group_28,  baseline_va_r10) %>% 
  group_by(severity_score, race_ethnicity) %>% 
  mutate(count_severity_race_ins_va = sum(n)) %>% 
  mutate(prop = n / count_severity_race_ins_va) %>% 
  ungroup() %>% 
  select(severity_score, race_ethnicity, vegf_group_28, baseline_va_r10, prop, count = n)


```


```{r}

# Graphing likelihood of Eylea as first treatment by severity, race, and private insurance
drug_sev_race_va %>% 
  # Asian filtered due to too small data size
  filter(!race_ethnicity %in% c("Unknown", "Other", "Asian")) %>% 
  filter(vegf_group_28 == "Eylea") %>% 
  filter(count > 25) %>%
  filter(baseline_va_r10 == 60) %>% 
  ggplot(aes(x = severity_score, y = prop, color = race_ethnicity)) +
  geom_point() + 
  geom_line() +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Likelihood of receiving Eylea at higher severity score by race",
    y = "Percentage of people receiving Eylea"
  ) 

```




#### Severity vs. vision
 
Do severity scores differ by race for people with the same vision score? 

```{r}

#calculate severity by race and age

severity_race_vision <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, baseline_va_letter) %>%
  summarize(avg_severity = mean(severity_score), count = n()) %>% 
  filter(count > 100) %>% 
  arrange(desc(avg_severity))

severity_race_vision

```

```{r}
# graph severity by race and age group

severity_race_vision %>%
  filter(race_ethnicity %in% c("Caucasian", "Black or African American", "Hispanic", "Asian")) %>% 
  ggplot(aes(x = baseline_va_letter, y = avg_severity, color = race_ethnicity)) +
  geom_line() +
  geom_point() +
  theme_light() +
  labs(
    title = "Severity by race and baseline VA", 
    y = "Average severity score", 
    x = "Baseline VA"
  )
```

Hispanic people are more likely to have a higher starting severity score at all levels of baseline visual acuity, as compared to Black, Asian, and Caucasian people. 


```{r}

#calculate severity by race and age

severity_race_vision_flip <-
  timeseries_analysis %>% 
  group_by(race_ethnicity, severity_score) %>%
  summarize(avg_va = mean(baseline_va_letter), count = n()) %>% 
  filter(count > 100) %>% 
  arrange(desc(avg_va))


severity_race_vision_flip %>%
  filter(race_ethnicity %in% c("Caucasian", "Black or African American", "Hispanic")) %>% 
  ggplot(aes(x = severity_score, y = avg_va, color = race_ethnicity)) +
  geom_line() +
  geom_point() +
  theme_light() +
  labs(
    title = "Severity by race and baseline VA", 
    x = "Severity score", 
    y = "Average Baseline VA"
  )

```





## Regression
#### Linear regression

Start with a linear regression to see what variables are initially predictive

```{r}
# basic linear regression
# one_year_va_delta_1.lm <- 
#   lm(
#     one_year_va_delta ~ relevel(race_ethnicity, ref = "Caucasian") + insurance + gender + first_dr_age + smoke_status + baseline_va_letter, 
#     data = timeseries_analysis %>% mutate(race_ethnicity = factor(race_ethnicity))
#     )
# 
# summary(one_year_va_delta_1.lm)
```

```{r}
# linear regression with insurance race interaction
# one_year_va_delta_2.lm <- 
#   lm(
#     one_year_va_delta ~ relevel(race_ethnicity, ref = "Caucasian") + insurance + relevel(race_ethnicity, ref = "Caucasian") * insurance + gender + first_dr_age + smoke_status + baseline_va_letter, 
#     data = timeseries_analysis %>% 
#       filter(insurance %in% c("Medicare", "Medicaid", "Private")) %>% 
#       mutate(race_ethnicity = factor(race_ethnicity))
#   )
# 
# summary(one_year_va_delta_2.lm)
```

```{r}
# linear regression with severity score
# one_year_va_delta_3.lm <- 
#   lm(one_year_va_delta ~ race_ethnicity + insurance + gender + first_dr_age + smoke_status + baseline_va_letter + severity_score, 
#      data = timeseries_analysis)
# 
# summary(one_year_va_delta_3.lm)
```

```{r}
# linear regression with race insurance interaction and severity score
# one_year_va_delta_4.lm <- 
#   lm(one_year_va_delta ~ race_ethnicity + insurance + race_ethnicity * insurance + gender + first_dr_age + smoke_status + baseline_va_letter + severity_score, 
#      data = timeseries_analysis %>% 
#        filter(insurance %in% c("Medicare", "Medicaid", "Private")) %>% 
#        mutate(
#          race_ethnicity = factor(race_ethnicity),
#          race_ethnicity = relevel(race_ethnicity, ref = "Caucasian"),
#          insurance = factor(insurance), 
#          insurance = relevel(insurance, ref = "Private")
#        )
#   ) 
# 
# summary(one_year_va_delta_4.lm)

```

#### Quantile regressionn

```{r}

```

